---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"
---

# Solutions: Rates of change, limits, and derivatives problem set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```

::: {.callout-tip}
## Instructions
These are the solutions to the [rates of change, limits, and derivatives problem set](day4-problem-set.qmd). Click a solution to expand it and try each problem yourself first!
:::


## Slope as a rate of change

**1)** Researchers plotted [dissolved oxygen (DO)](https://www.epa.gov/system/files/documents/2021-07/parameter-factsheet_do.pdf) concentration against water depth in a lake, and found the relationship was linear over the depth range sampled.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
depth <- seq(0, 12, length.out = 100)
DO <- -0.6*depth + 9

plot(depth, DO, type = "l", lwd = 3, col = "green",
     xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Depth (m)", ylab = "Dissolved oxygen (mg/L)",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 10, by = 1), col = "lightgray", lty = "dotted")
abline(v = seq(0, 12, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 12, by = 1))
axis(2, at = seq(0, 10, by = 1))
lines(depth, DO, lwd = 2)
```

a) Write the equation for dissolved oxygen, $DO$, as a function of depth.
b) Interpret the slope in a full sentence, including units.

::: {.callout-note collapse="true" title="Solution"}
**a)** The line passes through $(0, 9)$ and $(10, 3)$, so:

- $y$-intercept: $9$
- slope: $\frac{3-9}{10-0} = \frac{-6}{10} = -0.6$

$DO(depth) = -0.6 \cdot depth + 9$

**b)**

"For every additional meter of depth, dissolved oxygen concentration decreases by an average of 0.6 mg/L, starting from a surface concentration of about 9 mg/L."
:::



**2)** The growth of sharks can be modeled using [von Bertalanffy curves](https://en.wikipedia.org/wiki/Von_Bertalanffy_function). Scalloped hammerhead sharks in the Atlantic were found to have growth curves that follow the graph below.

```{r}
#| echo: false

vb<-function(a,l,k,t0){
  l*(1-exp(-k*(a-t0)))
}

x=seq(0,30)

length=vb(x,214.8,.13,-1.62)

p1<-ggplot()+
  geom_point(aes(x=x,y=length),color="black",size=3)+
  geom_line(aes(x=x,y=length),color="black",size=2)+
  theme_bw()+
  scale_y_continuous(breaks=seq(35,215,5))+
  scale_x_continuous(breaks=seq(0,32,by=2))+
  labs(x="Age (years)",y="Length (cm)")

p1
```

a) What is the average slope from ages 0 to 8? From 20 to 30?
b) What are the units of the average slopes and why would they change for a shark as they get older?
c) How would you find the instantaneous slope for a shark at age 12? What is the closest approximation you can make with this data and graph?

::: {.callout-note collapse="true" title="Solution"}
**a)**

The average slope from ages 0 to 8 can be calcualted by:
$$ slope = \frac{155-40}{8-0} = 14.375$$

The average slope from ages 0 to 8 can be calcualted by:
$$ slope = \frac{210-200}{30-20} =1 $$

This tells us that the average rate of change for sharks between the ages of 0 to 8 was 14.375. For the ages, 20 to 30 the average change was 1.

**b)**

Slopes are in units of length over age, so the slope indicates the growth in centimeters every year. As animals get older, ther slow down their growth. Less energy is spent in growing and more invested in reproductive energy.

<!-- **c)**

We would need to apply calculus to find the true answer, but we can approximate it reasonably well by finding the smallest possible difference between two points around age 12. In this case we can use $\frac{175-172}{13-12}$ to get an approximate instaneous rate of growth of 3 cm per year. -->
:::

![Scalloped hammerhead shark (Sphyrna lewini) off Costa Rica. [By Kris-Mikael Krister.](https://en.wikipedia.org/wiki/Scalloped_hammerhead#/media/File:Scalloped_hammerhead_shark_(Sphyrna_lewini)_Costa_Rica_(cropped).jpg) CC BY 2.0 ](images/Scalloped_hammerhead_shark_(Sphyrna_lewini)_Costa_Rica_(cropped).jpg){width=50%}