---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"  
---

# Rates of change, limits, and derivatives problem set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```

::: {.callout-tip}
## Instructions

Attempt each problem on your own before checking the [solutions](day4-problem-set-solutions.qmd).
Feel free to work with anyone in the cohort after giving the problems a try!
:::

## Slope as a rate of change

**1)** Researchers plotted [dissolved oxygen (DO)](https://www.epa.gov/system/files/documents/2021-07/parameter-factsheet_do.pdf) concentration against water depth in a lake, and found the relationship was linear over the depth range sampled.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
depth <- seq(0, 12, length.out = 100)
DO <- -0.6*depth + 9

plot(depth, DO, type = "l", lwd = 3, col = "green",
     xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Depth (m)", ylab = "Dissolved oxygen (mg/L)",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 10, by = 1), col = "lightgray", lty = "dotted")
abline(v = seq(0, 12, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 12, by = 1), cex.axis = 2)
axis(2, at = seq(0, 10, by = 1), cex.axis = 2)
lines(depth, DO, lwd = 2)
```

a) Write the equation for dissolved oxygen, $DO$, as a function of depth.
b) Interpret the slope in a full sentence, including units.

**2)** The growth of sharks can be modeled using [von Bertalanffy curves](https://en.wikipedia.org/wiki/Von_Bertalanffy_function). Scalloped hammerhead sharks in the Atlantic were found to have growth curves that follow the graph below.

```{r}
#| echo: false

vb<-function(a,l,k,t0){
  l*(1-exp(-k*(a-t0)))
}

x=seq(0,30)

length=vb(x,214.8,.13,-1.62)

p1<-ggplot()+
  geom_point(aes(x=x,y=length),color="black",size=3)+
  geom_line(aes(x=x,y=length),color="black",size=2)+
  theme_bw()+
  scale_y_continuous(breaks=seq(35,215,5))+
  scale_x_continuous(breaks=seq(0,32,by=2))+
  labs(x="Age (years)",y="Length (cm)")

p1
```

a) What is the average slope from ages 0 to 8? From 20 to 30? Include units.
b) What does the average slope represent in this scenario and why would these slopes change for a shark as they get older? Write you answers in complete sentences.
<!-- c) How would you find the instantaneous slope for a shark at age 12? What is the closest approximation you can make with this data and graph? -->

![Scalloped hammerhead shark (Sphyrna lewini) off Costa Rica. [By Kris-Mikael Krister.](https://en.wikipedia.org/wiki/Scalloped_hammerhead#/media/File:Scalloped_hammerhead_shark_(Sphyrna_lewini)_Costa_Rica_(cropped).jpg) CC BY 2.0 ](images/Scalloped_hammerhead_shark_(Sphyrna_lewini)_Costa_Rica_(cropped).jpg){width=50%}

## Limits and derivatives

**3)** Investigate what is the value of 
$$
\lim_{x\to 3}2^x.
$$


**4)** Let $f(x) = 2x^2+4$. Calculate $f'(x)$ and $f'(-3)$.

<br>

**5)** Calculate the derivative of $g(x) = x^4-5x^3+x-5$ at $x=5$.

<br>

**6)** What is the instantaneous rate of change of $h(x) = \frac{5}{x^2}$ at $x=2$?

<br>

**7)** 
  a. Approximate the slope of the tangent line in the graph below at $x=2$. 
  
  b. Write a short paragraph describing how the slope of the tangent line represents the derivative. 

```{r}
#| echo: false
#| code-summary: "Show the code"
x=seq(-4,4,by=0.1)
y=0.5*x^2-1


ggplot()+
  geom_line(aes(x=x,y=y),color="black",linewidth=3)+
  #annotate("segment",x=0,xend=4,y=-3,yend=5,color="red",linewidth=2)+
  theme_minimal()+
  scale_x_continuous(breaks=seq(-4,4,by=1))+
  scale_y_continuous(breaks=seq(-3,7,by=1))+
  theme(text=element_text(size=28))+
  labs(x="",y="")
  
```

## Comparing instantaneous and average rates of change

**8)** Since 2010, a mountain glacier has been losing volume at an accelerating pace as regional temperatures rise. Researchers model the cumulative ice volume lost, $V(t)$ (in km³), since monitoring began, as:

$$
V(t) = 0.5t^2 + 2t
$$

where $t$ is years since 2010.

(a) Use differentiation rules to find $V'(t)$.
(b) Evaluate $V'(5)$. What does this value mean in the context of the glacier? State it as a full sentence, with units.
(c) Now calculate the *average* rate of ice loss between $t=4$ and $t=8$.
(d) Are the values from (b) and (c) the same? In your own words, explain conceptually why an "average" rate of change over an interval doesn't have to match the "instantaneous" rate of change at a single point within it.
