---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"  
---

# Integrals and differential equations problem set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```

::: {.callout-tip}
## Instructions

Attempt each problem on your own before checking the [solutions](day5-problem-set-solutions.qmd).
Feel free to work with anyone in the cohort after giving the problems a try!
:::

## Exercise 6

After restoring a section of coastal wetland, ecologists measure the rate at which the wetland absorbs stormwater runoff, $r(t)$ (in m³/hour), throughout a 24-hour storm. The graph below shows this absorption rate over time. Partway through the storm, the monitoring sensor lost power and went offline between hour 10 and hour 14 (shaded region in the graph) — but you still need to report an estimate of the total volume of stormwater absorbed over the full 24-hour storm.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
t <- seq(0, 24, length.out = 300)
rate <- 200 * exp(-((t-9)^2)/40)

before <- t <= 10
after  <- t >= 14

plot(NA, xlim = c(0,24), ylim = c(0,220),
     xlab = "Hours since storm began", ylab = "Stormwater absorption rate (m³/hr)")
abline(v = seq(0,24, by = 6), col = "lightgray", lty = "dotted")

rect(10, 0, 14, 220, col = adjustcolor("gray", alpha.f = 0.25), border = NA)
text(12, 210, "sensor\noffline", cex = 0.8, col = "gray30")

lines(t[before], rate[before], lwd = 3, col = "#003660")
lines(t[after], rate[after], lwd = 3, col = "#003660")
points(10, rate[which.min(abs(t-10))], pch = 19, col = "#FEBC11", cex = 1.3)
points(14, rate[which.min(abs(t-14))], pch = 19, col = "#FEBC11", cex = 1.3)
```

(a) What does the *area under this curve* represent, in words? Be sure to include units.
(b) Looking at the shape of the graph (without calculating anything), during which 6-hour window — hours 0–6, 6–12, 12–18, or 18–24 — do you expect the *largest volume* of stormwater to have been absorbed? Explain your reasoning.
(c) Describe how you would handle the sensor gap when estimating the total area under the curve. What assumption(s) would you have to make, and how would the location of this particular gap affect your confidence in the final estimate?
