---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"
---
# $e^x$, Logs, and Lines Problem Set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```

::: {.callout-tip}
## Instructions

Attempt each problem on your own before checking the [solutions](day3-problem-set-solutions.qmd).
Feel free to work with anyone in the cohort after giving the problems a try!
:::


## $e^x$ and logarithms

**1)** Forestry officials discover an invasion of an invasive insect in a county park. They estimate the insect population grows exponentially, following $N(t) = N_0 e^{rt}$, where $N_0$ is the initial number of egg masses counted per hectare, and $t$ is time in years since detection.

At detection ($t=0$), officials count $N_0 = 150$ egg masses per hectare. Four years later, the count has grown to $400$ egg masses per hectare.

a) Solve for the growth rate $r$.
b) Forestry regulators will impose a mandatory quarantine zone if the population reaches $5{,}000$ egg masses per hectare. Using your value of $r$, solve for the time $t$ at which this threshold will be reached.
c) State your answer to (b) as a full sentence.

## Lines and linear functions

**2)** Find the equation of each of the following lines:


```{r, echo=FALSE}

x=seq(-5,5)

y=2*x+5

u=-3*x+2

p1<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=y),size=3,color="black")+
  geom_line(aes(x=x,y=y),size=2,color="black")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-5,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

p4<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=u),size=3,color="darkgreen")+
  geom_line(aes(x=x,y=u),size=2,color="darkgreen")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-15,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

cowplot::plot_grid(p1,p4,labels=c("A","D"))

```


**3)** A stream's water temperature increases as you move downstream from a shaded headwater spring. At a monitoring station 2 km downstream, the water temperature is 14°C. At a station 5 km downstream, the temperature is 20°C. Assume temperature increases linearly with distance downstream.

a) Find the equation for water temperature $T$ (°C) as a function of distance downstream $x$ (km).
b) Create a plot showing the line. Clearly label the axes, including units, and add a title.
c) What is the $y$-intercept of this line, and what does it mean in the context of the stream? Answer in a full sentence.

<br>

**4)** A regional market for albacore tuna is described below, where $Q$ is quantity in thousands of pounds sold per week and $P$ is price in \$ per pound.

(1) A fishery's weekly supply curve for tuna passes through $(Q,P) = (0, 20)$ and $(100, 70)$. Find the supply curve $S(Q)$ in slope-intercept form.

(2) At a price of \$70/lb, weekly quantity demanded is 40 (thousand lbs); at \$20/lb, quantity demanded is 140 (thousand lbs). Find the inverse demand equation $D(Q)$.

(3) Plot $S(Q)$ and $D(Q)$ on the same axes, with quantity on the $x$-axis and price on the $y$-axis. Label both axes clearly, including units.

(4) Find the price and quantity at which supply equals demand and interpret this point in one sentence.


---


*All the functions and example data are synthetic and were generated (some with the aid of Claude Code) for the purpose of this lesson.*