---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"
---

# Solutions: $e^x$, Logs, and Lines Problem Set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```


::: {.callout-tip}
## Instructions
These are the solutions to the [$e^x$, logs, and lines problem set](day3-problem-set.qmd). Click a solution to expand it and try each problem yourself first!
:::

## $e^x$ and logarithms

**1)** Forestry officials discover an invasion of an invasive insect in a county park. They estimate the insect population grows exponentially, following $N(t) = N_0 e^{rt}$, where $N_0$ is the initial number of egg masses counted per hectare, and $t$ is time in years since detection.

At detection ($t=0$), officials count $N_0 = 150$ egg masses per hectare. Four years later, the count has grown to $400$ egg masses per hectare.

a) Solve for the growth rate $r$.
b) Forestry regulators will impose a mandatory quarantine zone if the population reaches $5{,}000$ egg masses per hectare. Using your value of $r$, solve for the time $t$ at which this threshold will be reached.
c) State your answer to (b) as a full sentence.

::: {.callout-note collapse="true" title="Solution"}
**a)**
$$
\begin{aligned}
400 &= 150e^{4r} \\
\frac{400}{150} &= e^{4r} \\
\ln\!\left(\frac{400}{150}\right) &= 4r \\
r &= \frac{\ln(2.667)}{4} \approx 0.245
\end{aligned}
$$

**b)**
$$
\begin{aligned}
5000 &= 150e^{0.245t} \\
\frac{5000}{150} &= e^{0.245t} \\
\ln(33.33) &= 0.245t \\
t &= \frac{\ln(33.33)}{0.245} \approx 14.3
\end{aligned}
$$

**c)**

If the current growth trend continues, the insect population is expected to reach the quarantine threshold of 5,000 egg masses per hectare about 14.3 years after it was first detected.
:::

## Lines and linear functions

**2)** Find the slope and $y$-intercept of the following lines.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#|
x=seq(-5,5)

y=2*x+5

u=-3*x+2

p1<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=y),size=3,color="black")+
  geom_line(aes(x=x,y=y),size=2,color="black")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-5,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

p4<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=u),size=3,color="darkgreen")+
  geom_line(aes(x=x,y=u),size=2,color="darkgreen")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-15,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

cowplot::plot_grid(p1,p4,labels=c("A","D"))

```

::: {.callout-note collapse="true" title="Solution"}
The slope for graph A is 2 with a $y$-intercept of 5.

The slope for graph D is -3 with a $y$-intercept of 2
:::

**3)** A stream's water temperature increases as you move downstream from a shaded headwater spring. At a monitoring station 2 km downstream, the water temperature is 14°C. At a station 5 km downstream, the temperature is 20°C. Assume temperature increases linearly with distance downstream.

a) Find the equation for water temperature $T$ (°C) as a function of distance downstream $x$ (km).
b) Create a plot showing the line. Clearly label the axes, including units, and add a title.
c) What is the $y$-intercept of this line, and what does it mean in the context of the stream? Answer in a full sentence.

::: {.callout-note collapse="true" title="Solution"}
**a)** Using the two points $(2, 14)$ and $(5, 20)$:

$$
m = \frac{20-14}{5-2} = \frac{6}{3} = 2
$$

Using point-slope form with $(2,14)$:
$$
T - 14 = 2(x-2) \ \Rightarrow \ T = 2x + 10
$$

**b)**

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
x <- seq(0, 8, length.out = 100)
T <- 2*x + 10

plot(x, T, type = "l", lwd = 3, col = "green",
     xlim = c(0, 8), ylim = c(0, 24),
     xlab = "Distance downstream (km)", ylab = "Water temperature (°C)",
     main = "Stream temperature vs. distance downstream",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 24, by = 2), col = "lightgray", lty = "dotted")
abline(v = seq(0, 8, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 8, by = 1))
axis(2, at = seq(0, 24, by = 2))
points(c(2,5), c(14,20), pch = 19, cex = 1.3, col = "blue")
lines(x, T, lwd = 2)
```

**c)** The $y$-intercept is $b = 10$.

"According to this model, the water at the very start of the stream (right at the headwater spring, 0 km downstream) is predicted to be 10°C."
:::

**4)** A regional market for albacore tuna is described below, where $Q$ is quantity in thousands of pounds sold per week and $P$ is price in \$ per pound.

(1) A fishery's weekly supply curve for tuna passes through $(Q,P) = (0, 20)$ and $(100, 70)$. Find the supply curve $S(Q)$ in slope-intercept form.

(2) At a price of \$70/lb, weekly quantity demanded is 40 (thousand lbs); at \$20/lb, quantity demanded is 140 (thousand lbs). Find the inverse demand equation $D(Q)$.

(3) Plot $S(Q)$ and $D(Q)$ on the same axes, with quantity on the $x$-axis and price on the $y$-axis. Label both axes clearly, including units.

(4) Find the price and quantity at which supply equals demand and interpret this point in one sentence.



::: {.callout-note collapse="true" title="Solution"}

(1) Supply passes through $(0,20)$ and $(100,70)$:

$$
\begin{aligned}
m &= \frac{70-20}{100-0} = 0.5 &\text{Compute the slope} \\
S(Q) &= 20 + 0.5Q &\text{Use the } y\text{-intercept } (0,20)
\end{aligned}
$$

(2) Demand passes through $(Q,P)=(40,70)$ and $(140,20)$:

$$
\begin{aligned}
m &= \frac{20-70}{140-40} = -0.5 &\text{Compute the slope} \\
P &= -0.5Q + c \Rightarrow 70 = -0.5(40) + c \Rightarrow c =90 &\text{Using } (40,70) \\
D(Q) &= 90 - 0.5Q &\text{Simplify}
\end{aligned}
$$

(4) Set $S(Q) = D(Q)$:

$$
\begin{aligned}
20 + 0.5Q &= 90 - 0.5Q &\text{Set supply equal to demand} \\
Q &= 70 &\text{Solve for } Q \\
P &= 20 + 0.5(70) = 55 &\text{Plug into either curve}
\end{aligned}
$$

At a price of \$55/lb, fishers supply exactly 70 (thousand lbs/week) of tuna and buyers demand exactly that same amount.
:::