---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"  
---
# $e^x$, Logs, and Lines Problem Set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```

::: {.callout-tip}
## Instructions

Attempt each problem on your own before checking the [solutions](day3-problem-set-solutions.qmd).
Feel free to work with anyone in the cohort after giving the problems a try!
:::


## $e^x$ and logarithms

**1)** Forestry officials discover an invasion of an invasive insect in a county park. They estimate the insect population grows exponentially, following $N(t) = N_0 e^{rt}$, where $N_0$ is the initial number of egg masses counted per hectare, and $t$ is time in years since detection.

At detection ($t=0$), officials count $N_0 = 150$ egg masses per hectare. Four years later, the count has grown to $400$ egg masses per hectare.

a) Solve for the growth rate $r$.
b) Forestry regulators will impose a mandatory quarantine zone if the population reaches $5{,}000$ egg masses per hectare. Using your value of $r$, solve for the time $t$ at which this threshold will be reached.
c) State your answer to (b) as a full sentence.

## Lines and linear functions

**2)** Find the equation of each of the following lines:


```{r, echo=FALSE}

x=seq(-5,5)

y=2*x+5

u=-3*x+2

p1<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=y),size=3,color="black")+
  geom_line(aes(x=x,y=y),size=2,color="black")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-5,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

p4<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=u),size=3,color="darkgreen")+
  geom_line(aes(x=x,y=u),size=2,color="darkgreen")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-15,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

cowplot::plot_grid(p1,p4,labels=c("A","D"))

```


**3)** A stream's water temperature increases as you move downstream from a shaded headwater spring. At a monitoring station 2 km downstream, the water temperature is 14°C. At a station 5 km downstream, the temperature is 20°C. Assume temperature increases linearly with distance downstream.

a) Find the equation for water temperature $T$ (°C) as a function of distance downstream $x$ (km).
b) Create a plot showing the line. Clearly label the axes, including units, and add a title.
c) What is the $y$-intercept of this line, and what does it mean in the context of the stream? Answer in a full sentence.

## Slope as a rate of change

**4)** Researchers plotted [dissolved oxygen (DO)](https://www.epa.gov/system/files/documents/2021-07/parameter-factsheet_do.pdf) concentration against water depth in a lake, and found the relationship was linear over the depth range sampled.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
depth <- seq(0, 12, length.out = 100)
DO <- -0.6*depth + 9

plot(depth, DO, type = "l", lwd = 3, col = "green",
     xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Depth (m)", ylab = "Dissolved oxygen (mg/L)",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 10, by = 1), col = "lightgray", lty = "dotted")
abline(v = seq(0, 12, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 12, by = 1), cex.axis = 2)
axis(2, at = seq(0, 10, by = 1), cex.axis = 2)
lines(depth, DO, lwd = 2)
```

a) Write the equation for dissolved oxygen, $DO$, as a function of depth.
b) Interpret the slope in a full sentence, including units.

**5)** The growth of sharks can be modeled using [von Bertalanffy curves](https://en.wikipedia.org/wiki/Von_Bertalanffy_function). Scalloped hammerhead sharks in the Atlantic were found to have growth curves that follow the graph below.

```{r}

#| echo: false

vb<-function(a,l,k,t0){
  l*(1-exp(-k*(a-t0)))
}

x=seq(0,30)

length=vb(x,214.8,.13,-1.62)

p1<-ggplot()+
  geom_point(aes(x=x,y=length),color="black",size=3)+
  geom_line(aes(x=x,y=length),color="black",size=2)+
  theme_bw()+
  scale_y_continuous(breaks=seq(35,215,5))+
  scale_x_continuous(breaks=seq(0,32,by=2))+
  labs(x="Age (years)",y="Length (cm)")

p1
```

a) What is the average slope from ages 0 to 8? From 20 to 30? Include units.
b) What does the average slope represent in this scenario and why would these slopes change for a shark as they get older? Write you answers in complete sentences.
<!-- c) How would you find the instantaneous slope for a shark at age 12? What is the closest approximation you can make with this data and graph? -->

![Scalloped hammerhead shark (Sphyrna lewini) off Costa Rica. [By Kris-Mikael Krister.](https://en.wikipedia.org/wiki/Scalloped_hammerhead#/media/File:Scalloped_hammerhead_shark_(Sphyrna_lewini)_Costa_Rica_(cropped).jpg) CC BY 2.0 ](images/Scalloped_hammerhead_shark_(Sphyrna_lewini)_Costa_Rica_(cropped).jpg)

---


*All the functions and example data are synthetic and were generated (some with the aid of Claude Code) for the purpose of this lesson.*