---
sidebar: daily-materials
execute:
  freeze: true
author: "Carmen Galaz García, Nathaniel Grimes"
---
# Solutions: $e^x$, Logs, and Lines Problem Set

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,message=FALSE,warning=FALSE)

library(tidyverse)
```

::: {.callout-tip}
## Instructions
These are the solutions to the [$e^x$, logs, and lines problem set](day3-problem-set.qmd). Click a solution to expand it and try each problem yourself first!
:::

## $e^x$ and logarithms

**1)** Forestry officials discover an invasion of an invasive insect in a county park. They estimate the insect population grows exponentially, following $N(t) = N_0 e^{rt}$, where $N_0$ is the initial number of egg masses counted per hectare, and $t$ is time in years since detection.

At detection ($t=0$), officials count $N_0 = 150$ egg masses per hectare. Four years later, the count has grown to $400$ egg masses per hectare.

a) Solve for the growth rate $r$.
b) Forestry regulators will impose a mandatory quarantine zone if the population reaches $5{,}000$ egg masses per hectare. Using your value of $r$, solve for the time $t$ at which this threshold will be reached.
c) State your answer to (b) as a full sentence.

::: {.callout-note collapse="true" title="Solution"}
**a)**
$$
\begin{aligned}
400 &= 150e^{4r} \\
\frac{400}{150} &= e^{4r} \\
\ln\!\left(\frac{400}{150}\right) &= 4r \\
r &= \frac{\ln(2.667)}{4} \approx 0.245
\end{aligned}
$$

**b)**
$$
\begin{aligned}
5000 &= 150e^{0.245t} \\
\frac{5000}{150} &= e^{0.245t} \\
\ln(33.33) &= 0.245t \\
t &= \frac{\ln(33.33)}{0.245} \approx 14.3
\end{aligned}
$$

**c)**

If the current growth trend continues, the insect population is expected to reach the quarantine threshold of 5,000 egg masses per hectare about 14.3 years after it was first detected.
:::

## Lines and linear functions

**2)** Find the slope and $y$-intercept of the following lines.

```{r}
#| code-fold: true
#| code-summary: "Show the code"
#|
x=seq(-5,5)

y=2*x+5

u=-3*x+2

p1<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=y),size=3,color="black")+
  geom_line(aes(x=x,y=y),size=2,color="black")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-5,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

p4<-ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_point(aes(x=x,y=u),size=3,color="darkgreen")+
  geom_line(aes(x=x,y=u),size=2,color="darkgreen")+
  theme_bw()+
  theme(axis.line = element_blank())+
  scale_y_continuous(breaks=seq(-15,15,2))+
  scale_x_continuous(breaks=seq(-5,5,by=1))+
  labs(x="",y="")

cowplot::plot_grid(p1,p4,labels=c("A","D"))

```

::: {.callout-note collapse="true" title="Solution"}
The slope for graph A is 2 with a $y$-intercept of 5.

The slope for graph D is -3 with a $y$-intercept of 2
:::

**3)** A stream's water temperature increases as you move downstream from a shaded headwater spring. At a monitoring station 2 km downstream, the water temperature is 14°C. At a station 5 km downstream, the temperature is 20°C. Assume temperature increases linearly with distance downstream.

a) Find the equation for water temperature $T$ (°C) as a function of distance downstream $x$ (km).
b) Create a plot showing the line. Clearly label the axes, including units, and add a title.
c) What is the $y$-intercept of this line, and what does it mean in the context of the stream? Answer in a full sentence.

::: {.callout-note collapse="true" title="Solution"}
**a)** Using the two points $(2, 14)$ and $(5, 20)$:

$$
m = \frac{20-14}{5-2} = \frac{6}{3} = 2
$$

Using point-slope form with $(2,14)$:
$$
T - 14 = 2(x-2) \ \Rightarrow \ T = 2x + 10
$$

**b)**

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
x <- seq(0, 8, length.out = 100)
T <- 2*x + 10

plot(x, T, type = "l", lwd = 3, col = "green",
     xlim = c(0, 8), ylim = c(0, 24),
     xlab = "Distance downstream (km)", ylab = "Water temperature (°C)",
     main = "Stream temperature vs. distance downstream",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 24, by = 2), col = "lightgray", lty = "dotted")
abline(v = seq(0, 8, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 8, by = 1))
axis(2, at = seq(0, 24, by = 2))
points(c(2,5), c(14,20), pch = 19, cex = 1.3, col = "blue")
lines(x, T, lwd = 2)
```

**c)** The $y$-intercept is $b = 10$.

"According to this model, the water at the very start of the stream (right at the headwater spring, 0 km downstream) is predicted to be 10°C."
:::

## Slope as a rate of change

**4)** Researchers plotted [dissolved oxygen (DO)](https://www.epa.gov/system/files/documents/2021-07/parameter-factsheet_do.pdf) concentration against water depth in a lake, and found the relationship was linear over the depth range sampled.

```{r}
#| echo: false
#| fig-align: center
#| out-width: "70%"
depth <- seq(0, 12, length.out = 100)
DO <- -0.6*depth + 9

plot(depth, DO, type = "l", lwd = 3, col = "green",
     xlim = c(0, 12), ylim = c(0, 10),
     xlab = "Depth (m)", ylab = "Dissolved oxygen (mg/L)",
     xaxt = "n", yaxt = "n")
abline(h = seq(0, 10, by = 1), col = "lightgray", lty = "dotted")
abline(v = seq(0, 12, by = 1), col = "lightgray", lty = "dotted")
axis(1, at = seq(0, 12, by = 1))
axis(2, at = seq(0, 10, by = 1))
lines(depth, DO, lwd = 2)
```

a) Write the equation for dissolved oxygen, $DO$, as a function of depth.
b) Interpret the slope in a full sentence, including units.

::: {.callout-note collapse="true" title="Solution"}
**a)** The line passes through $(0, 9)$ and $(10, 3)$, so:

- $y$-intercept: $9$
- slope: $\frac{3-9}{10-0} = \frac{-6}{10} = -0.6$

$DO(depth) = -0.6 \cdot depth + 9$

**b)**

"For every additional meter of depth, dissolved oxygen concentration decreases by an average of 0.6 mg/L, starting from a surface concentration of about 9 mg/L."
:::



**5)** The growth of sharks can be modeled using [von Bertalanffy curves](https://en.wikipedia.org/wiki/Von_Bertalanffy_function). Scalloped hammerhead sharks in the Atlantic were found to have growth curves that follow the graph below.

```{r}
#| code-fold: true
#| code-summary: "Show the code"

vb<-function(a,l,k,t0){
  l*(1-exp(-k*(a-t0)))
}

x=seq(0,30)

length=vb(x,214.8,.13,-1.62)

p1<-ggplot()+
  geom_point(aes(x=x,y=length),color="black",size=3)+
  geom_line(aes(x=x,y=length),color="black",size=2)+
  theme_bw()+
  scale_y_continuous(breaks=seq(35,215,5))+
  scale_x_continuous(breaks=seq(0,32,by=2))+
  labs(x="Age (years)",y="Length (cm)")

p1
```

a) What is the average slope from ages 0 to 8? From 20 to 30?
b) What are the units of the average slopes and why would they change for a shark as they get older?
c) How would you find the instantaneous slope for a shark at age 12? What is the closest approximation you can make with this data and graph?

::: {.callout-note collapse="true" title="Solution"}
**a)**

The average slope from ages 0 to 8 can be calcualted by:
$$ slope = \frac{155-40}{8-0} = 14.375$$

The average slope from ages 0 to 8 can be calcualted by:
$$ slope = \frac{210-200}{30-20} =1 $$

This tells us that the average rate of change for sharks between the ages of 0 to 8 was 14.375. For the ages, 20 to 30 the average change was 1.

**b)**

Slopes are in units of length over age, so the slope indicates the growth in centimeters every year. As animals get older, ther slow down their growth. Less energy is spent in growing and more invested in reproductive energy.

<!-- **c)**

We would need to apply calculus to find the true answer, but we can approximate it reasonably well by finding the smallest possible difference between two points around age 12. In this case we can use $\frac{175-172}{13-12}$ to get an approximate instaneous rate of growth of 3 cm per year. -->
:::
