---
author: "Nathaniel Grimes"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE,warning=FALSE,message=FALSE)

library(tidyverse)
```

*Answer the following questions to the best of your ability.  Attempt exercises on your own first to make sure you fully understand the concepts. Feel free to work with anyone in the cohort after giving the problems a try!*

# Solutions: Derivatives Problem Set

## Derivative Rules

1) Explain how to find the derivative of the following functions. Be sure to explicitly denote which derivative rules (constant, multiple, sum, product, chain, etc.) you would use. For extra practice, have a try at solving them!

a) $\large f(x)=\sqrt{(x+3)}$
    
<span style="color:blue">
We can use chain rule to compute this one, with the outer function being $\sqrt(u)$ and ithe inner function being $x+3$. We can compute each of these derivatives using the power rule. We obtain that<\span>

<span style="color:blue">
$$
f'(x)=\frac{1}{2\sqrt{(x+3)}}
$$
</span>

    
b) $h(a)=\frac{\exp^{-a}}{a^2-5}$
    
<span style="color:blue">
First we would use the quotient rule then to solve for the individual derivatives. The individual derivatives will require us to use the chain rule (for the numerator) and the power rule (for the denominator). We obtain that
</span>

<span style="color:blue">
$$
h'(a)=\frac{-(a^2-5)e^{-a}-(2a)e^{-a}}{(a^2-5)^2}
$$
</span>

c) $g(t)=\left(\frac{5t^3+2}{2(t^2-2)}\right)^6$
    
<span style="color:blue">
First chain rule. In the $u$-substitution we have to use the quotient and power rules.
</span>

<span style="color:blue">
\begin{align}
g(t)&=\left(\frac{5t^3+2}{2(t^2-2)}\right)^6 \\
g(t)&=u(t)^6  &u=\frac{5t^3+2}{2(t^2-2)}\\
g'(t)&=6(u(t))^5\times u'(t)  &\frac{du}{dt}=\frac{2(t^2-2)15t^2-4t(5t^3+2)}{(2(t^2-2))^2} \\
g'(t)&=6(u(t))^5* \frac{2(t^2-2)15t^2-4t(5t^3+2)}{(2(t^2-2))^2} \\
g'(t)&=6\left(\frac{5t^3+2}{2(t^2-2)}\right)^5\frac{2(t^2-2)15t^2-4t(5t^3+2)}{(2(t^2-2))^2}
\end{align}
</span>
    
2) On the graph, label all regions where the following statements are true:

a) $f'(x)=0$
    
<span style="color:blue">
We have that $'f(x)=0$ at $x=-2$ and $x=2$, these are a minima and maxima respectively.
</span>
    
b) $f'(x)>0$
<span style="color:blue">    
The function $f(x)$ has a positive derivative whenever it is increasing, this happens over the interval (-2,0). This is indicated as the blue interval on the graph below.
</span>
    
c) $f'(x)<0$

<span style="color:blue">
The function $f(x)$ has a negative derivative whenever it is decreasing, this happens over the intervals $(-\infty,-2)$ and $(0,\infty)$. This range is indicated by the red intervals on the graph below.
</span>
    
    
d)$f''(x)>0$

<span style="color:blue">    
The second derivative $f''(x)$ is positive whenever the function is concave up (it has a U shape). The points at which the graph changes concavity are called inflection points. We can estimate the function is concave up whenever $x<-1$.
</span>
    
e) $f''(x)<0$
    
<span style="color:blue">    
The second derivative $f''(x)$ is negative whenever the function is concave down (it has an upside-down U shape). The points at which the graph changes concavity are called inflection points. We can estimate the function is concave down whenever $x>-1$.
</span>
    
```{r,fig.cap="The color bars correspond to the verbal statements and their colors in answer key above"}
x=seq(-3,1,by=.1)

y=-x^3-3*x^2+2

ggplot()+
  geom_hline(yintercept=0)+
  geom_vline(xintercept = 0)+
  geom_line(aes(x=x,y=y),color="black",size=3)+
  theme_classic()+
  labs(x="",y="")+
  annotate("segment",x=-3,xend=-2,y=0.1,yend=0.1,color="red",linewidth=2)+
  annotate("segment",x=0,xend=2,y=0.1,yend=0.1,color="red",linewidth=2)+
  annotate("segment",x=-2,xend=0,y=0.1,yend=0.1,color="blue",linewidth=2)
```
    

3)  Find the derivative of these equations

a) $f(x)=\exp^{x^2+3}$

\begin{align}
f(x)&=e^{x^2+3}\\
f'(x)&=2x e^{x^2+3}
\end{align}
    
b) $f(x)=\sqrt(2x)((x^3-4x+15))$

\begin{align}
f(x)&=\sqrt{(2x)}((x^3-4x+15)) \\
f'(x)&=\frac{1}{\sqrt{2x}}(x^3-4x+15)+\sqrt{2x}(3x^2-4)
\end{align}
    
c) $f(x)=\frac{x^3+2x}{x-1}$

\begin{align}
f(x)&=\frac{x^3+2x}{x-1}\\
f'(x)&=\frac{(x-1)(3x^2+2)-(x^3+2x)(1)}{(x-1)^2}
\end{align}
    
4) Find the critical points of the function and determine whether they are maximum or minimum.

$$
2x^3-5x^2+3x-12
$$

\begin{align}
f'(x)&=6x^2-10x+3=0 &\textcolor{blue}{\text{Use the Quadratic formula to solve}}\\
x&\approx 1.27  &x\approx 0.39 \\
f''(x)&=12x-10 \\
f''(1.27)&>0  &f''(0.39)&<0
\end{align}

\textcolor{blue}{Based on the first and second order conditions, there are two optimas for this function at $x\approx 1.27$ and $x\approx 0.39$. Plugging into the second order conditions we can see that $x\approx 0.39$ is a maximum and $x\approx 1.27$ is a minimum.}

    
5) Let's solve the canonical Gordon-Schaefer fisheries production model. This model formed the basis of fisheries management for half a century and is still a widely used model. It starts by defining the change in the abundance or biomass of fish by the logistic growth equation:

\begin{equation}
\frac{dB}{dt}=rB(t)(1-\frac{B(t)}{K})
\end{equation}

where $B$ is biomass, $t$ is time, $r$ is the intrinsic growth rate, and $K$ is the carrying capacity. Next, fishers catch fish based on their effort $E$, technological ability $q$ (also called the catchability coefficient), and the amount of fish $B$. For simplicity, we use an easy multiplicative form:

$$
Y=qEB
$$

This catch or harvest rate takes away from the growth of the fish stock, thus we can subtract it from the logistic growth equation 1. 

\begin{equation}
\frac{dB}{dt}=rB(t)(1-\frac{B(t)}{K})-Y
\end{equation}

**a)** Fishery managers are interested in finding the equilibrium level of biomass and harvest. We can set $\frac{dB}{dt}=0$. Solve for the equilibrium biomass in terms of K,q,r,E.

\textcolor{blue}{Solution}

\begin{align}
0=rB(1-\frac{B}{K})-qEB \\
qEB=rB(1-\frac{B}{K})\\
qE=r(1-\frac{B}{K})\\
\frac{rB}{K}=r-qE \\
B=K(1-\frac{qE}{r})
\end{align}

**b)** Sub the equilibrium biomass you found in A into the harvest equation. 

\textcolor{blue}{Solution}

\begin{align}
Y=qEB \\
Y=qEK(1-\frac{qE}{r})
\end{align}

**c)** What effort leads to a maximum sustainable yield? Maximize the equation from b) with respect to effort $E$.

\textcolor{blue}{We can use either the product rule or distribute the E. It's easier algebra in this case to distribute the terms then take the derivative}

\begin{align}
Y=qEK-\frac{q^2E^2K}{r}\\
\frac{dY}{dE}=qK-\frac{2q^2EK}{r}=0 \\
qK=\frac{2q^2EK}{r}\\
\frac{r}{2q}=E^*
\end{align}

**d)** Interpret what you found in c). What are the two variables that drive fisher harvest levels?

\textcolor{blue}{The amount of effort that leads to the most amount of catch is driven entirely by the rate of growth of fish stocks and the ability of fishers to catch them. If $q$ catchability goes up, fishers become more efficient and need less effort to reach max harvest. Faster growing species allow for more growth and thus more effort to catch them.}

**e)** Calculate what the maximum sustainable yield by substituting the optimal harvest level into the harvest equation.

\begin{align}
Y_{MSY}=\frac{qrK}{2q}-\frac{q^2(\frac{r}{2q})^2K}{r}\\
Y_{MSY}=\frac{rK}{2}-\frac{rK}{4}\\
Y_{MSY}=\frac{Kr}{4}
\end{align}